---
title: "Pseudotime_Monocle3"
output: html_document
---
# Sources
Tutorial video: https://www.youtube.com/watch?v=iq4T_uzMFcY&t=918s 
Data link: http://scrna.sklehabc.com/
Paper of this data: https://academic.oup.com/nsr/article/8/3/nwaa180/5896476?login=false 
Monocle3: https://cole-trapnell-lab.github.io/monocle3/docs/introduction/ 


### Installing Monocle 3

Install Bioconductor
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install(version = "3.14")
```

Impurt BC
```{r}
library(BiocManager)
```

Install dependencies that are not auto installed
```{r}
BiocManager::install(c('BiocGenerics', 'DelayedArray', 'DelayedMatrixStats',
                       'limma', 'lme4', 'S4Vectors', 'SingleCellExperiment',
                       'SummarizedExperiment', 'batchelor', 'HDF5Array',
                       'terra', 'ggrastr'))
```

Packages for Monocle3
```{r}
install.packages("remotes")
install.packages("R.utils")
install.packages("devtools")
devtools::install_github('cole-trapnell-lab/monocle3')
remotes::install_github('satijalab/seurat-wrappers')
```


Import other packages
```{r}
library(monocle3)
library(Seurat)
library(tidyverse)
library(SeuratWrappers)
library(remotes)
library(R.utils)
```

Import data
```{r}
markers <- read.delim('data/bcell/ABC_Marker.txt', header = T) # gene metadata
metadata <- read.delim('data/bcell/ABC_Meta.txt', header = T) # cell metadata
expr <- read.delim('data/bcell/ABC_umi_matrix_7551_cells.csv', header = T, sep = ',') # expression matrix
```

Create seurat object with countdata
```{r}
expr.t <- t(expr) # First transpose the data 
seu.obj <- CreateSeuratObject(counts = expr.t)

rownames(metadata) <- metadata$cell_id #convert cell_id to rowname

seu.obj <- AddMetaData(seu.obj, metadata = metadata) # Add the metadata 

validObject(seu.obj) # Check the object
```

Count mitochondrial percentage
```{r}
seu.obj$mitopercent <- PercentageFeatureSet(seu.obj, pattern = '^MT-')
```

Filter out low quality data
```{r}
seu.obj.filtered <- subset(seu.obj, subset = nCount_RNA > 800 &
                    nFeature_RNA > 500 &
                    mitopercent < 10)
```

subset my seurat object - B cells
```{r}
unique(seu.obj.filtered@meta.data$population)

Idents(seu.obj.filtered) <- seu.obj.filtered$population
b.seu <- subset(seu.obj.filtered, idents = "b")
b.seu
unique(b.seu@meta.data$redefined_cluster)

```

pre-processing using Seurat
```{r}
b.seu <- NormalizeData(b.seu)
b.seu <- FindVariableFeatures(b.seu)
b.seu <- ScaleData(b.seu)
b.seu <- RunPCA(b.seu)
b.seu <- FindNeighbors(b.seu, dims = 1:30)
b.seu <- FindClusters(b.seu, resolution = 0.9)
b.seu <- RunUMAP(b.seu, dims = 1:30, n.neighbors = 50)
```

Create cluster using DimPlot function of Seurat
```{r}
a1 <- DimPlot(b.seu, reduction = 'umap', group.by = 'redefined_cluster', label = T) + NoLegend()
a2 <- DimPlot(b.seu, reduction = 'umap', group.by = 'seurat_clusters', label = T) + NoLegend()
a1|a2
```

# MONOCLE3 WORKFLOW 

monocle3 requires cell_data_set object
convert Seurat object to cell_data_set object for monocle3
```{r}
cds <- as.cell_data_set(b.seu)
```

Explore the data
```{r}
colData(cds) # to get cell metadata
counts(cds) # to get counts data
fData(cds) # to gene metdata
rownames(fData(cds))[1:10]
```

since it misses the gene_short_name column, let's add it
```{r}
fData(cds)$gene_short_name <- rownames(fData(cds))
```

assign paritions
```{r}
reacreate.partition <- c(rep(1,length(cds@colData@rownames)))
names(reacreate.partition) <- cds@colData@rownames
reacreate.partition <- as.factor(reacreate.partition)
```


```{r}
cds@clusters$UMAP$partitions <- reacreate.partition
```

Assign the cluster info 
```{r}
list_cluster <- b.seu@active.ident
cds@clusters$UMAP$clusters <- list_cluster


# Assign UMAP coordinate - cell embeddings

cds@int_colData@listData$reducedDims$UMAP <- b.seu@reductions$umap@cell.embeddings
```

```{r}
# plot

#cluster.before.trajectory
plot_cells(cds,
           color_cells_by = 'cluster',
           label_groups_by_cluster = FALSE,
           group_label_size = 5) + NoLegend()

#cluster.names
plot_cells(cds,
           color_cells_by = "redefined_cluster",
           label_groups_by_cluster = FALSE,
           group_label_size = 5) +
  scale_color_manual(values = c('red', 'blue', 'green', 'maroon', 'yellow', 'grey', 'cyan')) +
  NoLegend()

```

Learn trajectory graph 
```{r}
cds <- learn_graph(cds, use_partition = FALSE)

plot_cells(cds,
           color_cells_by = 'redefined_cluster',
           label_groups_by_cluster = FALSE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE)
```

Order the cells in pseudotime
```{r}
cds <- order_cells(cds, reduction_method = 'UMAP', root_cells = colnames(cds[,clusters(cds) == 5]))

plot_cells(cds,
           color_cells_by = 'pseudotime',
           label_groups_by_cluster = TRUE,
           label_branch_points = FALSE,
           label_roots = FALSE,
           label_leaves = FALSE,
            group_label_size = 5)

```
Particular genes in trajectory
```{r}
plot_cells(cds,
           genes = c('E2F2', 'STMN1', 'CD52'),
           label_cell_groups = FALSE,
           show_trajectory_graph = FALSE,
           min_expr = 3)
```

cells ordered by monocle3 pseudotime
```{r}
pseudotime(cds)
cds$monocle3_pseudotime <- pseudotime(cds)
data.pseudo <- as.data.frame(colData(cds))

ggplot(data.pseudo, aes(monocle3_pseudotime, reorder(redefined_cluster, monocle3_pseudotime, median), fill = redefined_cluster)) +
  geom_boxplot() +
  ylab("Cells")+
  xlab("Pseudotime")+
  theme_classic()+
  NoLegend()
```

Finding genes that change as a function of pseudotime
```{r}
deg_bcells <- graph_test(cds, neighbor_graph = 'principal_graph', cores = 4)

deg_bcells %>% 
  arrange(q_value) %>% 
  filter(status == 'OK') %>% 
  head()

FeaturePlot(b.seu, features = c('E2F2', 'STMN1', 'CD52'))
```

Visualizing pseudotime in seurat
```{r}
b.seu$pseudotime <- pseudotime(cds)
Idents(b.seu) <- b.seu$redefined_cluster
FeaturePlot(b.seu, features = "pseudotime", label = T, cols = c("green4", "red3"))
```

